---
title: "data-joiner"
author: "riley wares"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading/installing packages. 
```{r load packages}

#install.packages("pacman")
library(pacman)

pacman::p_load(tidyverse, here, cowplot)

```


Reading in clean datasets. 
```{r import datasets}

beetle_clean <- read.csv(here("data", "processed", "beetle_clean_intermediate.csv"))
littercn_clean <- read.csv(here("data", "processed", "littercn_clean_intermediate.csv"))
littermass_clean <- read.csv(here("data", "processed", "littermass_clean_intermediate.csv"))

```


Creating a column for beetle abundance by counting the number of beetle observations 
per species per day. 
```{r count beetles}

beetle_abundance_taxonID <- beetle_clean %>%
  group_by(plotID, collectDate, taxonID) %>%
  summarize(count = n())

```
Creating a column for beetle richness per month by counting the number of unique 
beetle species per month. 
```{r count taxa}

beetle_richness_taxonID <- beetle_abundance_taxonID %>%
  group_by(plotID, collectDate) %>%
  summarize(richness = n())

```

A note: Only two decimal places are approved for this data. We should treat anything less than 0.01 as 0.

Created an average littermass dataset with columns for year, month and average littermass per month. 
```{r average litter mass}

litter_avg_mass <- littermass_clean %>%
  mutate(monthCollected=month(collectDate), yearCollected=year(collectDate)) %>%
  group_by(plotID, yearCollected, monthCollected) %>%
  summarize(avgmass = mean(dryMass))

```

Created an average carbon and nitrogen dataset with columns for year, month and average nitrogen and carbon percents per month. 
```{r average c and n}

litter_avg_cn <- littercn_clean %>%
  mutate(monthCollected=month(collectDate), yearCollected=year(collectDate)) %>%
  group_by(plotID, yearCollected, monthCollected) %>%
  summarize(avgCpercent = mean(carbonPercent), avgNpercent = mean(nitrogenPercent))

```

Combined the average littermass and average nitrogen and carbon datasets to have one
litter data sets with all averages. 
```{r full join litter data}

litter_averages_total <- litter_avg_mass %>%
  full_join(litter_avg_cn)

```

```{r export data}

litter_averages_total %>%
  write_csv(here("data", "processed", "litter_averages_total.csv"))

beetle_abundance_taxonID %>%
  write_csv(here("data", "processed", "beetle_abundance_taxonID.csv"))

beetle_richness_taxonID %>%
  write_csv(here("data", "processed", "beetle_richness_taxonID.csv"))

```


DAY TWO WORK

```{r }
beetle_abundance_monthly <- beetle_abundance_taxonID %>% 
  mutate(monthCollected=month(collectDate), 
         yearCollected=year(collectDate)) 

```
