---
title: "megapit_data_wrangle"
author: "zachary krausman"
output: html_document
date: "2026-02-24"
---

Setup (do not change).
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in (and install if needed) tidyverse and here packages. 
```{r load packages, message = FALSE}

#install.packages("tidyverse")
#install.packages("here")

library(tidyverse)
library(here)

```


Read in all three csvs for the beetle data, litter chemistry data, and litter dry mass data.
``` {r read data}
beetles_raw <- read_csv(here("data", "raw", "bet_parataxonomistID.csv"))
litter_carbonnitrogenraw <- read_csv(here("data", "raw", "ltr_litterCarbonNitrogen.csv"))
litter_massraw <- read_csv(here("data", "raw", "ltr_massdata.csv"))

```

View all newly created raw data objects to see what we are working with.
``` {r view data}
view(beetles_raw)
view(litter_carbonnitrogenraw)
view(litter_massraw)
```


Selected only necessary columns from the raw data sets. 
``` {r collecting variables}
beetles <- beetles_raw %>%
  select(siteID, plotID, collectDate, taxonID, uid)

littercn <- litter_carbonnitrogenraw %>%
  select(siteID, plotID, collectDate, nitrogenPercent, carbonPercent, uid)

littermass <- litter_massraw %>%
  select(siteID, plotID, collectDate, dryMass, uid)
```

View data sets to check for NAs.
```{r view modified tables}
view(littercn)
view(littermass)
```


Removed NAs from the data sets to leave only "clean" data sets. 
```{r remove na data}
littercnclean <- littercn %>%
  filter(!is.na(carbonPercent),
         !is.na(nitrogenPercent))

littermassclean <- littermass %>%
  filter(!is.na(dryMass))

beetleclean <- beetles %>%
  filter(!is.na(taxonID))

view(littercnclean) 
view(littermassclean) 
view(beetleclean)
```

Write csvs for processed data and save to processed data folder. 
```{r intermediate data processing}
littercnclean %>%
  write_csv(here("data", "processed", "littercn_clean_intermediate.csv"))
littermassclean %>%
  write_csv(here("data", "processed", "littermass_clean_intermediate.csv"))
beetleclean %>%
  write_csv(here("data", "processed", "beetle_clean_intermediate.csv"))

```






