---
title: "megapit_data_wrangle"
author: "zachary krausman"
output: html_document
date: "2026-02-24"
---

Setup (do not change).
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in (and install if needed) tidyverse and here packages. 
```{r load packages, message = FALSE}

#install.packages("tidyverse")
#install.packages("here")

library(tidyverse)
library(here)

```


Read in all three csvs for the beetle data, litter chemistry data, and litter dry mass data.
``` {r read data}

beetles_raw <- read_csv(here("data", "raw", "bet_parataxonomistID.csv"))

litter_carbonnitrogenraw <- read_csv(here("data", "raw", "ltr_litterCarbonNitrogen.csv"))

litter_massraw <- read_csv(here("data", "raw", "ltr_massdata.csv"))

```

View all newly created raw data objects to see what we are working with.
``` {r view data}

view(beetles_raw)
view(litter_carbonnitrogenraw)
view(litter_massraw)

```

Selected only necessary columns from the raw data sets. 
``` {r collecting variables}

beetles <- beetles_raw %>%
  select(siteID, plotID, collectDate, taxonID, uid)

littercn <- litter_carbonnitrogenraw %>%
  select(siteID, plotID, collectDate, nitrogenPercent, carbonPercent, uid)

littermass <- litter_massraw %>%
  select(siteID, plotID, collectDate, dryMass, uid)

```

We collected data from multiple domain sites (D01, D02, and D07). As a note from NEON via DP1.10111.001 (Site management and event reporting), the data was "obfuscated to a higher taxonomic rank when the taxon was found to be listed as threatened, endangered, or sensitive at the state level where the observation was recorded". This was removed from data submitted for all sites except D01 and D20. We are using some sites from D01, but since many arthropods are not listed with protections, it is possible that some taxon levels are obfuscated, and some are not (i.e. some beetles maybe be listed at a higher taxonomic level instead of a genus or species resolution).

View data sets to check for NAs.
```{r view modified tables}

view(littercn)
view(littermass)

```

Removed NAs from the data sets to leave only "clean" data sets. 
```{r remove na data}

littercnclean <- littercn %>%
  filter(!is.na(carbonPercent),
         !is.na(nitrogenPercent))

littermassclean <- littermass %>%
  filter(!is.na(dryMass))

beetleclean <- beetles %>%
  filter(!is.na(taxonID))

view(littercnclean) 
view(littermassclean) 
view(beetleclean)

```

Write csvs for processed data and save to processed data folder. 
```{r intermediate data processing}

littercnclean %>%
  write_csv(here("data", "processed", "littercn_clean_intermediate.csv"))
littermassclean %>%
  write_csv(here("data", "processed", "littermass_clean_intermediate.csv"))

littermassclean %>%
  write_csv(here("data", "processed", "littermass_clean_intermediate.csv"))

beetleclean %>%
  write_csv(here("data", "processed", "beetle_clean_intermediate.csv"))

```
